Challenges in Developing Great Quasi-Monte Carlo Software

### 2023-03-07 MCQMC2022ChallengesQMCSoftwareArticle.tex
Intermediate changes

In the next section we describe why QMC software is important.  We then discuss characteristics of great QMC software:

@@

QMC software makes theoretical advances in QMC available to practitioners. However, translating pseudo-code into good executable code is nuanced.  Software developers must write code that is computationally efficient, numerical stable, and provides reasonable default choices of tuning parameters. Great QMC software relieves users of these concerns.

@@

Great QMC software opens up new application areas for QMC methods by allowing practitioners to compare new methods to existing ones.  QMC software has been successful in quantitative finance, uncertainty quantification, and image rendering. Unexpectedly good or bad computational results leads to open theoretical questions.  For example, the early application of QMC to high dimensional integrals arising in financial risk \cite{PasTra95} led to a wave of theoretical results on the tractability of integration in weighted spaces \cite{Woz99a,DicEtal14a,NovWoz10a}.

We next discuss characteristics of great QMC software outlined in the introduction.  For each characteristic we identify what is lacking and what might be done to remedy the lack.

@@

Expanding on the problem formulation in the introduction, we want to approximate well the integral or expectation, $\mu$, by the sample mean, $\hat{\mu}$:
\begin{gather}
\label{eq:cubSummary}
	\mu : = \int_\calT g(\bst) \, \lambda(\bst) \, \D \bst  = \bbE[f(\bsX)] = \int_{[0,1]^d} f(\bsx)  \, \D \bsx \approx \frac 1n \sum_{i=1}^{n} f(\bsx_i) =: \hat{\mu}_n, \\
 \label{CDHJS_eq:error_tol}
 \abs{\mu - \hat{\mu}_n} \le \varepsilon.
\end{gather}
Progressing from the original problem to a satisfactory solution requires several software components, which we describe in the subsections below.  Great QMC software environments allows different implementations of these components to be freely interchanged.

@@

Lattice and digital sequences are not unique, and improved or alternative versions continue to be constructed.  Joe and Kuo \cite{JoeKuo03,JoeKuo08,SobolDirection} have proposed  Sobol' generator direction numbers, which are now widely used, but were not part of the early LD sequence software.  L'Ecuyer and colleagues have created LatNetBuilder \cite{LatNet} to construct good lattice and polynomial lattice sequences based on user defined criteria. As was raised during a lunch at MCQMC 2022, we should agree on a consistent format for storing the parameters that define good LD sequences so that they can by shared across software libraries.

@@

Complex integrands might be best evaluated by libraries that have no direct connection to QMC software libraries.  In the uncertainty quantification (UQ) space, the recently developed UQ and Model Bridge (UM-Bridge) \cite{umbridge}, connects libraries that generate sequences of model parameters, $\bsx_1, \bsx_2, \ldots$ with libraries that evaluate models, $f(\bsx_1), f(\bsx_2), \ldots$.  We illustrate in Figure \ref{fig:beam} how UM-Bridge connects QMCPy with a differential equation boundary value problem solver for a cantilevered beam with random Young's modulus \cite{ParSee22a}, where the randomness is given by $\bsX \sim \calU[1,1.2]^3$.  The beam position is discretized at $31$ points.  QMCPy computes the expected deflection of the beam a these points for a sequence of error tolerances using lattice rules.  By taking advantage of virtual machines and docker containers, UM-Bridge provides automatic load balancing to facilitate seamless parallel model evaluation.% the evaluation of the beam deflection can be computed in parallel in multiple cores of a single CPU to decrease execution time.

@@

QMC software is sometimes written by those not intimately familiar with QMC.  Those of us most familiar with good QMC practice need to vigilantly be aware of QMC algorithms in popular software and inform the developers when their software contains bugs or deviates from best practice.

\section{Efficient} \label{CDHJS_sec:efficient}
Another obvious feature of great QMC software is numerical efficiency.  This requires developers to identify bottlenecks in performance.  Intensive computation may need to be rewritten in lower level languages, like C.  Code may need to be refactored to avoid excessive calls to memory.  As new hardware architectures arise, QMC software should adapt.

Efficiency gains in QMC software can also be the result clever algorithm choices.  The Bayesian stopping rules for lattice and Sobol sequences in \cite{RatHic19a,JagHic22a} choose covariance kernels that match the corresponding LD sequence.  This reduces the computations from what would be $\calO(n^3)$ for arbitrary covariance kernels to only $\calO(n \log n)$ for these well-chosen kernels.

>>>>

In the next section, we describe why QMC software is important.  We then discuss the characteristics of great QMC software:

@@

QMC software makes theoretical advances in QMC available to practitioners. However, translating pseudo-code into good executable code is nuanced.  Software developers must write code that is computationally efficient, numerically stable, and provides reasonable default choices of tuning parameters. Great QMC software relieves users of these concerns.

@@

Great QMC software opens up new application areas for QMC methods by allowing practitioners to compare new methods to existing ones.  QMC software has been successful in quantitative finance, uncertainty quantification, and image rendering. Unexpectedly good or bad computational results lead to open theoretical questions.  For example, the early application of QMC to high dimensional integrals arising in financial risk \cite{PasTra95} led to a wave of theoretical results on the tractability of integration in weighted spaces \cite{Woz99a,DicEtal14a,NovWoz10a}.

We next discuss the characteristics of great QMC software outlined in the introduction.  For each characteristic, we identify what is lacking and what might be done to remedy the lack.

@@

Expanding on the problem formulation in the introduction, we want to approximate well the integral or expectation, $\mu$, by the sample mean, $\hat{\mu}$:
\begin{gather}
\label{eq:cubSummary}
	\mu : = \int_\calT g(\bst) \, \lambda(\bst) \, \D \bst  = \bbE[f(\bsX)] = \int_{[0,1]^d} f(\bsx)  \, \D \bsx \approx \frac 1n \sum_{i=1}^{n} f(\bsx_i) =: \hat{\mu}_n, \\
 \label{CDHJS_eq:error_tol}
 \abs{\mu - \hat{\mu}_n} \le \varepsilon.
\end{gather}
Progressing from the original problem to a satisfactory solution requires several software components, which we describe in the subsections below.  Great QMC software environments allow different implementations of these components to be freely interchanged.

@@

Lattice and digital sequences are not unique, and improved or alternative versions continue to be constructed.  Joe and Kuo \cite{JoeKuo03,JoeKuo08,SobolDirection} have proposed  Sobol' generator direction numbers, which are now widely used, but were not part of the early LD sequence software.  L'Ecuyer and colleagues have created LatNetBuilder \cite{LatNet} to construct good lattice and polynomial lattice sequences based on user-defined criteria. As was raised during a lunch at MCQMC 2022, we should agree on a consistent format for storing the parameters that define good LD sequences so that they can be shared across software libraries.

@@

Complex integrands might be best evaluated by libraries that have no direct connection to QMC software libraries.  In the uncertainty quantification (UQ) space, the recently developed UQ and Model Bridge (UM-Bridge) \cite{umbridge}, connects libraries that generate sequences of model parameters, $\bsx_1, \bsx_2, \ldots$ with libraries that evaluate models, $f(\bsx_1), f(\bsx_2), \ldots$.  We illustrate in Figure \ref{fig:beam} how UM-Bridge connects QMCPy with a differential equation boundary value problem solver for a cantilevered beam with random Young's modulus \cite{ParSee22a}, where the randomness is given by $\bsX \sim \calU[1,1.2]^3$.  The beam position is discretized at $31$ points.  QMCPy computes the expected deflection of the beam at these points for a sequence of error tolerances using lattice rules.  By taking advantage of virtual machines and docker containers, UM-Bridge provides automatic load balancing to facilitate seamless parallel model evaluation.% the evaluation of the beam deflection can be computed in parallel in multiple cores of a single CPU to decrease execution time.

@@

QMC software is sometimes written by those not intimately familiar with QMC.  Those of us most familiar with good QMC practices need to vigilantly be aware of QMC algorithms in popular software and inform the developers when their software contains bugs or deviates from best practices.

\section{Efficient} \label{CDHJS_sec:efficient}
Another obvious feature of great QMC software is numerical efficiency.  This requires developers to identify bottlenecks in performance.  The intensive computations may need to be rewritten in lower-level languages, like C.  Code may need to be refactored to avoid excessive calls to memory.  As new hardware architectures arise, QMC software should adapt.

Efficiency gains in QMC software can also be the result of clever algorithm choices.  The Bayesian stopping rules for lattice and Sobol sequences in \cite{RatHic19a,JagHic22a} choose covariance kernels that match the corresponding LD sequence.  This reduces the computations from what would be $\calO(n^3)$ for arbitrary covariance kernels to only $\calO(n \log n)$ for these well-chosen kernels.
